Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-08-15 latex/main.tex

Input genome alignments are masked in sc2ts 
so that deletions and ambiguous bases are treated as missing data by the HMM, 
and once the samples are incorporated into the ARG, 
these bases are imputed to carry the values inherited by the node in question. 
Deletions are an important part of SARS-CoV-2 evolution [cites], however, 
and to incorporate them into the ARG we identified a set of [XXX] sites most frequently involved in deletions [cite source?] 
and post-hoc mapped alignment data for these sites to the ARG using parsimony. 
Although sites are treated independently in this approach, 
the approximation successfully identified the most frequent deletions, 
such as the recurrent 21991-21993 (see Document S1 for more details). 
The node representing the origin of Alpha, with 290,407 descendants, 
is correctly associated with key lineage-defining deletions: 
11288-11296 (ORF1A del3675-3677), 
21765-21770 (Spike del69-70), 
21991-21993 (Spike del144), and 
28271 (non-coding). 
The Spike 69-70 deletion, which causes S-Gene Target Failure (REF), 
also emerges twice in the origin of Omicron - 
in the node leading to the BA.1 variant (342,099 descendants), 
and the node leading to the BA.4 variant (189,746 descendants). 
To speed up inference and to avoid issues caused by sites enriched for errors [citation?], 
sc2ts masks out a set of 100 sites identified as the most highly homoplasic 
by running on a subset of the data (STAR Methods). 
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER [citation?]. 
In addition to remapping the data for sites identified as being involved in deletions, 
we added mutations for XX of these highly recurrent sites 
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony. 
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods). 
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment 
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata). 
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG, 
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).

>>>>

Input genome alignments are masked in sc2ts 
so that deletions and ambiguous bases are treated as missing data by the HMM, 
and once the samples are incorporated into the ARG, 
these bases are imputed to carry the values inherited by the node in question. 
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however, 
and to incorporate them into the ARG we identified a set of [XXX] sites 
most frequently involved in deletions [cite source?] 
and post-hoc mapped alignment data for these sites to the ARG using parsimony. 
Although sites are treated independently in this approach, 
the approximation successfully identified the most frequent deletions, 
such as the recurrent 21991-21993 (see Document S1 for more details). 
The node representing the origin of Alpha, with 290,407 descendants, 
is correctly associated with key lineage-defining deletions: 
11288-11296 (ORF1A del3675-3677), 
21765-21770 (Spike del69-70), 
21991-21993 (Spike del144), and 
28271 (non-coding). 
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021}, 
also emerges twice in the origin of Omicron - 
in the node leading to the BA.1 variant (342,099 descendants), 
and the node leading to the BA.4 variant (189,746 descendants). 
To speed up inference and to avoid issues caused by sites enriched for errors [citation?], 
sc2ts masks out a set of 100 sites identified as the most highly homoplasic 
by running on a subset of the data (STAR Methods). 
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER [citation?]. 
In addition to remapping the data for sites identified as being involved in deletions, 
we added mutations for XX of these highly recurrent sites 
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony. 
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods). 
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment 
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata). 
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG, 
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).
